Trauma treatment delay

### 2023-12-12 manuscriptlongversion.Rmd
text

Karolinska University hospital, which is the equivalent of a trauma
level 1 hospital, admits 1500 trauma patients each year. To be added to
the trauma registry, a patient must be over 15 years of age, had ISS\>9
and/or had trauma code activation. They get added to both the Karolinska
Trauma registry as well as the national trauma registry (SweTrau). The
registry includes data on vital signs, injuries, interventions and
patient information. Another database, the trauma care quality database
includes information relevant for MoM conferences. At Karolinska
University Hospital, patients who die during hospital stay gets their
case sent to a MoM everytime. However, for the rest of the patients, two
specialized nurses determine whether they are. Several audit filters are
set up to flag for patients with possible OFIs. Audit filters may
include low SBP, high RR, EMS scene time greater than 20 mins etc.
Flagged patients are then reviewed by these two nurses who decide if
there is a possible OFI, in which case their case is sent to a MoM
conference for further review.

@@

We conducted a registry based cohort study using data from the trauma
registry and trauma care quality database at the Karolinska University
Hospital in Solna. The trauma registry includes about 14,022 patients
treated between 2012 and 2022. The trauma care quality database is a
subset of the trauma registry and includes about 8000 patients selected
for review between 2013 and 2022. This project will link the two
databases and assess how different patient level factors, such as age,
sex, blood pressure, and injury severity, are associated with delayed
treatment using logistic regression.

@@

For missing hospital values in GCS, RR and SBP, values from EMT
personnel were used. This was accomplished by identifying missing data
in the column for the hospital value of the variable and, if available,
replacing the row in the dataset with the prehospital value. If a
prehospital value is missing, the patient is excluded from the study.

@@

The data comes from the trauma registry and trauma care quality
database.

@@

Selection bias as this is a retrospective, pots hoc cohort study where
the outcome has occurred.

@@

Using the one in ten rule, 7 independent variables were used for the
logistical regression given that we have 72 patients with delayed
treatment. The one in ten rule states that for every variable examined,
10 events are needed in the group. In this case, this means that because
we have 72 patients with the examined outcome, we can have 7 variables.

The logistical regression is run through Rstudio using the glm command.
It was ran several times to make an ajusted table and 7 unadjusted, one
for each variable To make the tables, the package gtsummary was used to
visualize the adjusted and unadjusted regression.

@@

These results show that there seems to be a correlation between some
variables and the presence of delayed treatment. A lower GCS had a
correlation with delayed treatment while a higher ISS had the same
correlation. This suggests that a more heavily injured and less awake
patient is more prone to receiving delayed treatment. Compared to
previous a previous study analysing the same data but with ofi as the
outcome, this result is similar to what they found using the same
database [@Pat_factors]. As the mentioned study suggests, this could be
explained by that lower GCS indicates severe trauma, aswell as higher
ISS. More injured patients often entails a more complex management and
require more interventions. More interventions leave more room for delay
to these interventions, why they are more prone to delay in treatment.

Respiratory rate also had the correlation that lower respiratory rate
correlated with lower rates of delayed treatment among the patients. As
the interval is \>10 to \<29 in the triage guideline, both too low and
too high respiratory rate is not desirable[@EHCP2015]. This might have a
similar explanation to ISS and GCS, that bradypnea in a patient might
attract more attention in the hospital and lower the incidence of
delayed treatment. Another possible explanation might be that patients
with tachypnea might receive higher rates of delayed treatment because
they are more severely injured, therefore a higher risk of missing a
possible injury. Another similar study, which focused on ofi generally,
showed the same correlation, although the significance dissapeared in
the adjusted analysis [@Pat_factors].

@@

The fact that age and gender did not show an increased odds ratio should
be seen as a positive sign. It suggests that among trauma patients,
there is no systematical delay based on gender or age.

@@

Further discussing possible biases, misclassification bias might also be
present in this study, though it is something that we can't control in
this study. As previously mentioned, the possible existence of an ofi is
decided during the MoM conferences, where they then add it to the
database. However, they might not always be right in their judgement and
misclassify from time to time, leading to patients landing in the wrong
group in the analysis. Also, some patients might have been missed even
before reaching the MoM conference. The patients included in the
conference are selected through audit filters and two specialized
nurses, and patients might be missed in this stage. This bias should not
be big enough to have a significant impact on the results, but should be
kept in mind.

@@

Interpretation and clinical applications

Further discussing the results and discussing it's clinical
applications, the results can be interpreted as that it is mostly
injured and less awake patients that receive delayed. The clinical
applications for this are not clear, but caregivers should already be
aware that the more injuries a patients has, the more room for error
exists. It is good though to receive a reminder that it's important to
always work towards minimizing the delay for these injured patients.

Generalisability

>>>>

Karolinska University hospital, which is the equivalent of a trauma
level 1 hospital, admits 1500 trauma patients each year. To have your
information added to the trauma registry, a patient must be over 15
years of age, had ISS\>9 and/or had trauma code activation. They get
added to both the Karolinska Trauma registry as well as the national
trauma registry (SweTrau). The registry includes data on vital signs,
injuries, interventions and patient information. Another database, the
trauma care quality database includes information relevant for MoM
conferences. At Karolinska University Hospital, patients who die during
hospital stay gets their case sent to a MoM everytime. However, for the
rest of the patients, two specialized nurses determine whether they are.
Several audit filters are set up to flag for patients with possible
OFIs. Audit filters may include low SBP, high RR, EMS scene time greater
than 20 mins etc. Flagged patients are then reviewed by these two nurses
who decide if there is a possible OFI, in which case their case is sent
to a MoM conference for further review.

@@

We conducted a registry based cohort study using data from the trauma
registry and trauma care quality database at the Karolinska University
Hospital in Solna. The trauma registry includes about 14,022 patients
treated between 2012 and 2022. The trauma care quality database is a
subset of the trauma registry and includes about 8000 patients selected
for review between 2013 and 2022. This project linked the two databases
and assessed how different patient level factors, such as age, sex,
blood pressure, and injury severity, are associated with delayed
treatment using logistic regression.

@@

For missing hospital values in GCS, RR and SBP, values collected from
EMT personnel before arrival to hospital were used. This was
accomplished by identifying missing data in the column for the hospital
value of the variable and, if available, replacing the row in the
dataset with the prehospital value. If a prehospital value is missing,
the patient is excluded from the study.

@@

The data comes from the trauma registry and trauma care quality
database,

@@

Selection bias as this is a retrospective post hoc cohort study where
the outcome has occurred before the study began.

@@

Using the one in ten rule, 7 independent variables were used for the
logistical regression given that we have 70 patients with delayed
treatment. The one in ten rule states that for every variable examined,
10 events are needed in the group. In this case, this means that because
we have 70 patients with the examined outcome, we can have 7 variables.

The logistical regression is run through Rstudio using the glm command.
It was ran several times to make an ajusted table and 7 unadjusted, one
for each variable. To make the tables, the package gtsummary was used to
visualize the adjusted and unadjusted regression individually, then
merging them together into one table.

@@

These results show a possible correlation between some variables and the
presence of delayed treatment. A lower GCS had a correlation with
delayed treatment while a higher ISS had the same correlation. This
suggests that a more heavily injured and less awake patient is more
prone to receiving delayed treatment. Compared to previous a previous
study analysing the same data but with ofi as the outcome, this result
is similar to what they found using the same database [@Pat_factors]. As
the mentioned study suggests, this could be explained by that lower GCS
indicates severe trauma, aswell as higher ISS. More injured patients
often entails a more complex management and require more interventions.
More interventions leave more room for delay to these interventions, why
they are more prone to delay in treatment.

Respiratory rate also had the correlation that lower respiratory rate
correlated with lower rates of delayed treatment among the patients. As
the interval is \>10 to \<29 in the triage guideline, both too low and
too high respiratory rate is not desirable[@EHCP2015]. This might have a
similar explanation to ISS and GCS, that patients with tachypnea might
receive higher rates of delayed treatment because they are more severely
injured, therefore a higher risk of missing a possible injury. Another
possible explanation might be bradypnea in a patient might attract more
attention in the hospital and lower the incidence of delayed treatment.
Another similar study, which focused on ofi generally, showed the same
correlation, although the significance dissapeared in the adjusted
analysis [@Pat_factors].

@@

The fact that age and gender did not show an increased odds ratio should
be seen as a positive sign. It suggests that among trauma patients,
there is no systematic delay based on gender or age.

@@

On discussing possible biases, misclassification bias might also be
present in this study, though it is something that we can't control in
this study. As previously mentioned, the possible existence of an ofi is
decided during the MoM conferences, where they then add it to the
database. However, they might not always be right in their judgement and
misclassify from time to time, leading to patients landing in the wrong
group in the analysis. Also, some patients might have been missed even
before reaching the MoM conference. The patients included in the
conference are selected through audit filters and two specialized
nurses, and patients might be missed in this stage. This bias should not
be big enough to have a significant impact on the results, but can be
kept in mind.

@@

Further analysing the results and it's clinical applications, the
results can be interpreted as that it is mostly injured and less awake
patients that receive delayed. The clinical applications for this are
not clear, but caregivers should already be aware that the more injuries
a patients has, the more room for error exists. It is good though to
receive a reminder that it's important to always work towards minimizing
the delay for these injured patients.
